hbm_paper_results

### 2024-03-09 paper_replication_results.Rmd
methods & results

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(data.table, ggplot2, tidyverse, ggseg, ggsegExtra, paletteer, ggpubr, gridExtra, lemon, parallel, rstatix, lme4, interactions, ggrepel, ggrain)
#Sys.setenv("MC_CORES"=10L)
options("mc.cores")

@@

cent.sex.test.med %>%
  # dplyr::filter(dataset == "cf.gam" | dataset == "cf.gamlss") %>%
  ggplot() +
  geom_density(aes(x=median_of_median_sex_diff, fill=dataset, color=dataset), alpha=0.6, position = "identity") +
  geom_vline(xintercept = 0, linetype="dashed") +
  facet_wrap(~pheno_cat, scales="free")

t.test.cent.sex.test.med <- cent.sex.test.med %>%
  group_by(dataset, pheno_cat) %>%
  summarise(t.test.p = tidy(t.test(median_of_median_sex_diff))$p.value) %>%
  mutate(p.val_fdr = p.adjust(t.test.p, method="fdr", n = nrow(cent.sex.test.med)),
                  sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                                      p.val_fdr >= 0.05 ~ FALSE))
t.test.cent.sex.test.med %>%
  ggplot() +
  geom_col(aes(x=pheno_cat, y=sig_fdr, fill=dataset), position="dodge")
```

@@

# raw_mean_cent %>%
#   dplyr::filter(perm=="perm-001") %>%
#   dplyr::filter(VolReg_mean_centile > .85) %>% nrow()

@@

perm_subj.df2_long <- perm_subj.df2 %>%
  rename(allpheno_mean_centile = mean_centile) %>%
  pivot_longer(cols=contains("mean_centile"), names_to="pheno_cat", values_to="mean_centile") %>%
  mutate(pheno_cat = gsub("_mean_centile", "", pheno_cat))

high_cent.sum <- perm_subj.df2_long %>%
  dplyr::filter(mean_centile > 0.80) %>%
  group_by(dataset, perm, pheno_cat) %>%
  summarise(
    male_ratio = sum(sex == "Male") / n(),
    female_ratio = sum(sex == "Female") / n()
  )

high_cent.sum %>%
  #dplyr::filter(dataset != "raw") %>%
  ggplot(aes(x=dataset, y=female_ratio)) +
  geom_boxplot() +
  # geom_dotplot(binaxis = "y", stackdir="center") +
  facet_wrap(~pheno_cat, scales="free_y") +
  #geom_hline(yintercept=high_cent.sum[high_cent.sum$dataset == "raw" & high_cent.sum$perm=="perm-001",][["female_ratio"]], linetype="dashed")+
  labs(title="subjects with high average centiles")

low_cent.sum <- perm_subj.df2_long %>%
  dplyr::filter(mean_centile < 0.20) %>%
  group_by(dataset, perm, pheno_cat) %>%
  summarise(
    male_ratio = sum(sex == "Male") / n(),
    female_ratio = sum(sex == "Female") / n()
  )

low_cent.sum %>%
  #dplyr::filter(dataset != "raw") %>%
  ggplot(aes(x=dataset, y=female_ratio)) +
  geom_boxplot() +
  facet_wrap(~pheno_cat, scales="free_y") +
  # geom_dotplot(binaxis = "y", stackdir="center")  +
  #geom_hline(yintercept= low_cent.sum[low_cent.sum$dataset == "raw" & low_cent.sum$perm=="perm-001",][["female_ratio"]], linetype="dashed") +
  labs(title="subjects with low average centiles")

ext_cent.sum <- perm_subj.df2_long %>%
  dplyr::filter(mean_centile < 0.20 | mean_centile > 0.80) %>%
  group_by(dataset, perm, pheno_cat) %>%
  summarise(
    male_ratio = sum(sex == "Male") / n(),
    female_ratio = sum(sex == "Female") / n()
  )

ext_cent.sum %>%
  #dplyr::filter(dataset != "raw") %>%
  ggplot(aes(x=dataset, y=female_ratio)) +
  geom_boxplot() +
  # geom_dotplot(binaxis = "y", stackdir="center")  +
  #geom_hline(yintercept= ext_cent.sum[ext_cent.sum$dataset == "raw" & ext_cent.sum$perm=="perm-001",][["female_ratio"]], linetype="dashed")+
  facet_wrap(~pheno_cat, scales="free_y") +
  labs(title="subjects with extreme average centiles")

perm_subj.df %>%
  # dplyr::filter(dataset == "cf.gam" | dataset == "cf.gamlss") %>%
  ggplot() +
  geom_histogram(aes(x=mean_cent_diff, fill=sex, color=sex), position="identity", alpha=0.6, bins=60) +
  geom_vline(xintercept = 0, linetype="dashed") +
  facet_wrap(~dataset, scales="free")
```

linear ComBat -> too few women with high centiles (avg. across phenos, pattern varies), too many women with low centiles; ~ balances out overall (except regional volume)

ComBat-GAM -> too many women with high centiles (small but pretty consistent pattern), too many women with low centiles (patter varies a bit across phenos); OVERALL TOO MANY WOMEN W EXTREME CENTILES

@@

```{r}
#one-sample tests against raw vals

one_samp_wilcox <- function(cf_method, df){
  df2 <- df %>%
    dplyr::filter(dataset == cf_method)
  raw_val <- df[df$dataset == "raw" & df$perm=="perm-001",][["male_ratio"]]
  wilcox.test(df2$male_ratio, mu=raw_val)
}

cf_test_list <- unique(perm_subj.df$dataset)

high_test <- lapply(cf_test_list, one_samp_wilcox, df=high_cent.sum)
names(high_test) <- cf_test_list
high_test

low_test <- lapply(cf_test_list, one_samp_wilcox, df=low_cent.sum)
names(low_test) <- cf_test_list
low_test

ext_test <- lapply(cf_test_list, one_samp_wilcox, df=ext_cent.sum)
names(ext_test) <- cf_test_list
ext_test
```

@@

perm_subj.df2.by.site %>%
  ggplot() +
  geom_boxplot(aes(x=sim.site, y=mean_cent_abs.diff_site, fill=dataset)) +
  facet_wrap(~ dataset, scales="free")

@@

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(data.table, ggplot2, tidyverse, ggseg, ggsegExtra, paletteer, ggpubr, gridExtra, lemon, parallel, rstatix, lme4, interactions, ggrepel, ggrain, gdata)
#Sys.setenv("MC_CORES"=10L)
options("mc.cores")

@@

t.test.cent.sex.test.med <- cent.sex.test.med %>%
  group_by(dataset, pheno_cat) %>%
  summarise(t.test.p = tidy(t.test(median_of_median_sex_diff))$p.value,
            t.test.est = tidy(t.test(median_of_median_sex_diff))$estimate,
            x_pos = median(median_of_median_sex_diff)) %>% #test median of each distribution against 0
  mutate(p.val_fdr = p.adjust(t.test.p, method="fdr", n = length(unique(cent.sex.test.med$dataset))*length(unique(cent.sex.test.med$pheno_cat))),
                  sig_fdr = case_when(p.val_fdr < 0.001 ~ "***",
                               p.val_fdr >= 0.001 & p.val_fdr < 0.01 ~ "**",
                               p.val_fdr >= 0.01 & p.val_fdr < 0.05 ~ "*",
                               p.val_fdr >= 0.05 ~ "")) %>%
  dplyr::filter(dataset == "cf.gam" | dataset == "cf.gamlss")

t.test.cent.sex.test.med %>%
  ggplot() +
  geom_col(aes(x=pheno_cat, y=sig_fdr, fill=dataset), position="dodge")

sex_bias_density_plt <- cent.sex.test.med %>%
  dplyr::filter(dataset == "cf.gam" | dataset == "cf.gamlss") %>%
  ggplot() +
  geom_density(aes(x=median_of_median_sex_diff*100, fill=dataset, color=dataset), alpha=0.6, position = "identity") +
  geom_vline(xintercept = 0, linetype="dashed", color="#3F3F3F") +
  facet_wrap(~pheno_cat) +
  theme_linedraw() +
  geom_text(aes(label = sig_fdr, x= x_pos*100, y=ifelse(t.test.cent.sex.test.med$dataset == "cf.gam", 10.2, 5.5)),
              data = t.test.cent.sex.test.med) +
  scale_fill_manual(values = c(cf= "#440D54", cf.gam = "#3D75C2", cf.lm = "#75C23D", cf.gamlss ="#CFB023FF"), name = "ComBat \nConfiguration", labels=dataset_names) +
  scale_color_manual(values = c(cf= "#440D54", cf.gam = "#3D75C2", cf.lm = "#75C23D", cf.gamlss ="#CFB023FF"), name = "ComBat \nConfiguration", labels=dataset_names) +
  theme(legend.position = "bottom")+
  labs(x="Medians of Sex-Differences in Centile Errors", y="Density Across 100 Replications")

cent.sex.test.med %>%
  dplyr::filter(dataset == "cf.gam" | dataset == "cf.gamlss" | dataset == "cf.lm") %>%
  ggplot() +
  geom_boxplot(aes(y=median_of_median_sex_diff*100, x=dataset, color=dataset)) +
  geom_hline(yintercept = 0, linetype="dashed", color="#3F3F3F") +
  facet_wrap(~pheno_cat, scales="free") +
  theme_linedraw() +
  geom_text(aes(label = sig_fdr, x= x_pos*100, y=10.2),
              data = t.test.cent.sex.test.med) +
  scale_fill_manual(values = c(cf= "#440D54", cf.gam = "#3D75C2", cf.lm = "#75C23D", cf.gamlss ="#CFB023FF"), name = "ComBat \nConfiguration", labels=dataset_names) +
  scale_color_manual(values = c(cf= "#440D54", cf.gam = "#3D75C2", cf.lm = "#75C23D", cf.gamlss ="#CFB023FF"), name = "ComBat \nConfiguration", labels=dataset_names) +
  theme(legend.position = "bottom")+
  labs(x="Medians of Sex-Differences in Centile Errors Within Brain Feature Types", y="Density Across 100 Replications")
```

@@

raw_mean_cent %>%
  dplyr::filter(perm=="perm-001") %>%
  dplyr::filter(SA_mean_centile < .2) %>% 
  nrow()

@@

perm_subj.df2_long <- perm_subj.df2 %>%
  rename(allpheno_mean_centile = mean_centile) %>%
  pivot_longer(cols=contains("mean_centile"), names_to="pheno_cat", values_to="mean_centile") %>%
  mutate(pheno_cat = gsub("_mean_centile", "", pheno_cat)) %>%
  mutate(pheno_cat = case_when(pheno_cat == "VolGlob" ~ "Global Volume",
                               pheno_cat == "VolReg" ~ "Regional Volume",
                               pheno_cat == "CT" ~ "Regional CT",
                               pheno_cat == "SA" ~ "Regional SA"))

high_cent.sum <- perm_subj.df2_long %>%
  dplyr::filter(pheno_cat != "allpheno") %>%
  dplyr::filter(mean_centile > 0.80) %>%
  group_by(dataset, perm, pheno_cat) %>%
  summarise(
    male_ratio = sum(sex == "Male") / n(),
    female_ratio = sum(sex == "Female") / n()
  )

low_cent.sum <- perm_subj.df2_long %>%
  dplyr::filter(pheno_cat != "allpheno") %>%
  dplyr::filter(mean_centile < 0.20) %>%
  group_by(dataset, perm, pheno_cat) %>%
  summarise(
    male_ratio = sum(sex == "Male") / n(),
    female_ratio = sum(sex == "Female") / n()
  )

ext_cent.sum <- perm_subj.df2_long %>%
  dplyr::filter(pheno_cat != "allpheno") %>%
  dplyr::filter(mean_centile < 0.20 | mean_centile > 0.80) %>%
  group_by(dataset, perm, pheno_cat) %>%
  summarise(
    male_ratio = sum(sex == "Male") / n(),
    female_ratio = sum(sex == "Female") / n()
  )
```

```{r}
#get raw vals for horizontal lines
high_cent.sum_raw <- high_cent.sum %>%
  dplyr::filter(dataset == "raw" & perm=="perm-001")

low_cent.sum_raw <- low_cent.sum %>%
  dplyr::filter(dataset == "raw" & perm=="perm-001")

ext_cent.sum_raw <- ext_cent.sum %>%
  dplyr::filter(dataset == "raw" & perm=="perm-001")

#test significance
get_t_tests <- function(df, raw_df){
  #df <- df_list[[cat]]
  cat <- unique(df$pheno_cat)
  #print(cat)
  mu <- raw_df$female_ratio[which(raw_df$pheno_cat == cat)]
  #print(mu)
  out <- df %>%
    dplyr::filter(dataset != "raw") %>%
    group_by(dataset) %>%
    summarise(p_val = t.test(female_ratio, mu=mu)$p.val,
              pheno_cat = cat,
              max_val = max(female_ratio),
              mu=mu) %>%
    mutate(p.val_fdr = p.adjust(p_val, method="fdr", n = (4*4*3)), #adjusting for 4 methods * 4 pheno classes * 3 centile categories
           sig_fdr = case_when(p.val_fdr < 0.001 ~ "***",
                               p.val_fdr >= 0.001 & p.val_fdr < 0.01 ~ "**",
                               p.val_fdr >= 0.01 & p.val_fdr < 0.05 ~ "*",
                               p.val_fdr >= 0.05 ~ "NS"))
  return(out)
}

high_t_test.df <- lapply(split(high_cent.sum, high_cent.sum$pheno_cat), get_t_tests, raw_df=high_cent.sum_raw) %>% bind_rows()
low_t_test.df <- lapply(split(low_cent.sum, low_cent.sum$pheno_cat), get_t_tests, raw_df=low_cent.sum_raw) %>% bind_rows()
ext_t_test.df <- lapply(split(ext_cent.sum, ext_cent.sum$pheno_cat), get_t_tests, raw_df=ext_cent.sum_raw) %>% bind_rows()

#merge dfs
all_cent_sum.df <- gdata::combine(high_cent.sum, low_cent.sum, ext_cent.sum) %>%
  mutate(source = gsub("_cent.sum", "", source))
all_cent.sum_raw <- all_cent_sum.df %>%
  dplyr::filter(dataset == "raw" & perm=="perm-001")

all_t_test.df <- gdata::combine(high_t_test.df, low_t_test.df, ext_t_test.df) %>%
  mutate(source = gsub("_t_test.df", "", source))

#scale by raw value
all_cent_sum.df_scaled <- all_cent_sum.df %>%
  mutate(dataset = factor(dataset, levels=c("raw", "cf", "cf.lm", "cf.gam", "cf.gamlss"), ordered=TRUE)) %>%
  arrange(dataset) %>%
  group_by(source, pheno_cat) %>%
  mutate(female_ratio_sc = female_ratio/first(female_ratio),
         female_ratio_diff = female_ratio - first(female_ratio))

#plots!
all_cent_sum.df %>%
  dplyr::filter(dataset != "raw") %>%
  ggplot(aes(x=dataset, y=female_ratio)) +
  geom_boxplot(aes(color=dataset)) +
  # geom_dotplot(binaxis = "y", stackdir="center") +
  facet_grid(pheno_cat~source, scales="free_y") +
  geom_hline(data=all_cent.sum_raw, aes(yintercept=female_ratio), linetype="dashed")+
  geom_text(aes(label = sig_fdr, y = max_val + 0.01),
              data = all_t_test.df) +
  scale_color_manual(values = c(cf= "#440D54", cf.gam = "#3D75C2", cf.lm = "#75C23D", cf.gamlss ="#CFB023FF"),  name = "ComBat Configuration", labels=dataset_names) + 
  theme_linedraw()

cent_ext_list <- c(ext = ">80% and <20%", high = ">80%", low = "<20%")

all_cent_sum_boxplt <- all_cent_sum.df %>%
  dplyr::filter(dataset != "raw" & dataset != "cf") %>%
  mutate(source = factor(source, levels=c("low", "high", "ext"), ordered=TRUE)) %>%
  ggplot(aes(x=dataset, y=female_ratio)) +
  geom_boxplot(aes(color=dataset)) +
  # geom_dotplot(binaxis = "y", stackdir="center") +
  facet_grid(pheno_cat~source, scales="free_y", labeller = labeller(source=cent_ext_list)) +
  geom_hline(data=all_cent.sum_raw, aes(yintercept=female_ratio), linetype="dashed")+
  # geom_text(aes(label = sig_fdr, y = max_val + 0.01),
  #             data = all_t_test.df[which(all_t_test.df$dataset != "cf"),]) +
  scale_color_manual(values = c(cf= "#440D54", cf.gam = "#3D75C2", cf.lm = "#75C23D", cf.gamlss ="#CFB023FF")) + 
  theme_linedraw() +
  scale_x_discrete(labels=dataset_names) +
  # scale_y_continuous(labels=function(x)x/all_t_test.df$female_ratio[which(raw_df$pheno_cat == pheno_cat)]) + 
  theme(legend.position="none", plot.title = element_text(size=12), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(y="Proportion Female", x = "Dataset")

all_cent_sum.df_scaled %>%
  dplyr::filter(dataset != "raw" & dataset != "cf") %>%
  mutate(source = factor(source, levels=c("low", "high", "ext"), ordered=TRUE)) %>%
  ggplot(aes(x=dataset, y=female_ratio_sc)) +
  geom_boxplot(aes(color=dataset)) +
  # geom_dotplot(binaxis = "y", stackdir="center") +
  facet_grid(pheno_cat~source, scales="free_y", labeller = labeller(source=cent_ext_list)) +
  geom_hline(yintercept=1, linetype="dashed")+
  # geom_text(aes(label = sig_fdr, y = max_val + 0.01),
  #             data = all_t_test.df[which(all_t_test.df$dataset != "cf"),]) +
  scale_color_manual(values = c(cf= "#440D54", cf.gam = "#3D75C2", cf.lm = "#75C23D", cf.gamlss ="#CFB023FF")) + 
  theme_linedraw() +
  scale_x_discrete(labels=dataset_names) +
  # scale_y_continuous(labels=function(x)x/all_t_test.df$female_ratio[which(raw_df$pheno_cat == pheno_cat)]) + 
  theme(legend.position="none", plot.title = element_text(size=12), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(y="Female Bias", x = "Dataset")

all_cent_sum_boxplt.diff <- all_cent_sum.df_scaled %>%
  dplyr::filter(dataset != "raw" & dataset != "cf") %>%
  mutate(source = factor(source, levels=c("low", "high", "ext"), ordered=TRUE)) %>%
  ggplot(aes(x=dataset, y=female_ratio_diff)) +
  geom_boxplot(aes(color=dataset)) +
  # geom_dotplot(binaxis = "y", stackdir="center") +
  facet_grid(pheno_cat~source, scales="free_y", labeller = labeller(source=cent_ext_list)) +
  geom_hline(yintercept=0, linetype="dashed")+
  # geom_text(aes(label = sig_fdr, y = max_val + 0.01),
  #             data = all_t_test.df[which(all_t_test.df$dataset != "cf"),]) +
  scale_color_manual(values = c(cf= "#440D54", cf.gam = "#3D75C2", cf.lm = "#75C23D", cf.gamlss ="#CFB023FF")) + 
  theme_linedraw() +
  scale_x_discrete(labels=dataset_names) +
  # scale_y_continuous(labels=function(x)x/all_t_test.df$female_ratio[which(raw_df$pheno_cat == pheno_cat)]) + 
  theme(legend.position="none", plot.title = element_text(size=12), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(y="Female Bias", x = "Dataset")

all_cent_sum_boxplt_wtest <- all_cent_sum.df %>%
  dplyr::filter(dataset != "raw") %>%
  mutate(source = factor(source, levels=c("low", "high", "ext"), ordered=TRUE)) %>%
  ggplot(aes(x=dataset, y=female_ratio)) +
  geom_boxplot(aes(color=dataset)) +
  # geom_dotplot(binaxis = "y", stackdir="center") +
  facet_grid(pheno_cat~source, scales="free_y", labeller = labeller(source=cent_ext_list)) +
  geom_hline(data=all_cent.sum_raw, aes(yintercept=female_ratio), linetype="dashed")+
  geom_text(aes(label = sig_fdr, y = max_val + 0.01),
              data = all_t_test.df) +
  scale_color_manual(values = c(cf= "#440D54", cf.gam = "#3D75C2", cf.lm = "#75C23D", cf.gamlss ="#CFB023FF")) + 
  theme_linedraw() +
  scale_x_discrete(labels=dataset_names) +
  theme(legend.position="none", plot.title = element_text(size=12), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(y="Proportion Female", x = "Dataset")

all_cent_sum.df %>%
  ggplot(aes(x=dataset, y=female_ratio)) +
  geom_boxplot(aes(color=dataset)) +
  # geom_dotplot(binaxis = "y", stackdir="center") +
  facet_wrap(~source, scales="free_y") +
  #geom_hline(data=all_cent.sum_raw, aes(yintercept=female_ratio), linetype="dashed")+
  # geom_text(aes(label = sig_fdr, y = max_val + 0.01),
  #             data = all_t_test.df) +
  scale_color_manual(values = c(cf= "#440D54", cf.gam = "#3D75C2", cf.lm = "#75C23D", cf.gamlss ="#CFB023FF"),  name = "ComBat Configuration", labels=dataset_names) + 
  theme_linedraw()
```

linear ComBat ~ too few women with high centiles, too many women with low centiles; ~ balances out overall

ComBat-GAM -> variable pattern, pretty close to raw

@@

@@

@@

## FIGURE 3
```{r}
blank <- ggplot() + theme_void()
sex.var.plot.global <- readRDS("figs/sex_var_plot_global.RDS")
sex.var.plot <- readRDS("figs/sex_var_plot.RDS")

fig_3 <- ggarrange(
  ggarrange(
  ggarrange(blank, sex.var.plot.global, sex.var.plot,
  ncol=3, nrow=1, 
  widths = c(.1, .8, 3.1),
  hjust=0.01,
  align = "hv",
  common.legend = TRUE, legend = "bottom"
  ),
  sex_bias_density_plt,
  nrow=2, ncol=1,
  labels=c("A", "B")
  ),
  all_cent_sum_boxplt.diff,
  ncol=2, nrow=1,
  labels=c(NA, "C"),
  widths = c(1, 1)
  # #align = "hv",
  # heights=c(1, 1.8),
  # common.legend = FALSE,
  # #labels=c("A", "B")
  )

fig_3
ggsave("figs/figure_3.jpeg", fig_3, width=15, height=10, units="in")
```

@@

Figure 3 but with sig. tests for boxplots

```{r}
sex.var.plot.global <- readRDS("figs/sex_var_plot_global.RDS")
sex.var.plot <- readRDS("figs/sex_var_plot.RDS")

fig_3_wtest <- ggarrange(
  ggarrange(
  ggarrange(sex.var.plot.global, sex.var.plot,
  ncol=2, nrow=1, 
  widths = c(1, 3.25),
  hjust=0.01,
  align = "hv",
  common.legend = TRUE, legend = "bottom"
  ),
  sex_bias_density_plt,
  nrow=2, ncol=1,
  labels=c("A", "B")
  ),
  all_cent_sum_boxplt_wtest,
  ncol=2, nrow=1,
  labels=c(NA, "C"),
  widths = c(1, 1.8)
  # #align = "hv",
  # heights=c(1, 1.8),
  # common.legend = FALSE,
  # #labels=c("A", "B")
  )

fig_3_wtest
ggsave("figs/figure_3_wtest.jpeg", fig_3, width=15, height=10, units="in")
```
